\section{Related Work}
\label{sec:related_work}

\paragraph{MAP-Elites and Grid-Based QD.}
MAP-Elites~\cite{mouret2015illuminating} partitions the behavior space into a regular grid, maintaining one elite per cell.
Its simplicity has driven broad adoption across robotics~\cite{cully2015robots}, game design~\cite{gravina2019procedural}, and aerodynamic optimization~\cite{gaier2018data}.
CMA-ME~\cite{fontaine2020covariance} improves MAP-Elites by using CMA-ES emitters for directed search within niches.
CMA-MAE~\cite{fontaine2022differentiable} adds archive-awareness to the emitters.
However, all grid-based methods face exponential cell growth with behavior dimension, limiting practical use to $d \leq 5$.

\paragraph{Voronoi-Based QD.}
CVT-MAP-Elites~\cite{vassiliades2018using} replaces the regular grid with Voronoi tessellation using $k$-means centroids, providing more flexible partitioning.
This avoids some pathologies of uniform grids but still requires pre-specifying the number of niches and computing assignments to centroids, whose cost grows with dimension.
Multi-dimensional Archive of Phenotypic Elites (MAPE)~\cite{vassiliades2018using} extends this with adaptive centroid placement.

\paragraph{High-Dimensional QD.}
Recent work has begun addressing the dimensionality challenge.
PGA-MAP-Elites~\cite{nilsson2021policy} uses policy gradient updates within MAP-Elites for high-dimensional action spaces, but the behavior space itself remains low-dimensional.
Aurora~\cite{paolo2021sparse} learns a low-dimensional behavior embedding via autoencoders, sidestepping the problem by projecting to lower dimensions.
VQ-Elites~\cite{duarte2024vq} uses vector quantization for adaptive discretization.
These approaches modify the behavior representation; \mmr instead changes the archive maintenance mechanism to work directly in high-dimensional behavior spaces.

\paragraph{Maximum Marginal Relevance in IR.}
MMR was introduced by Carbonell and Goldstein~\cite{carbonell1998use} for diversified document retrieval and summarization.
It selects documents that balance query relevance with novelty relative to already-selected items.
MMR has been extended with learned similarity functions~\cite{xia2015learning}, applied to recommendation systems~\cite{agrawal2009diversifying}, and analyzed as submodular optimization~\cite{lin2011class}.
To our knowledge, \mmr is the first application of MMR to evolutionary optimization.

\paragraph{Submodular Optimization for Diversity.}
The connection between diversity maximization and submodular optimization is well-established~\cite{nemhauser1978analysis, krause2014submodular}.
The greedy algorithm achieves a $(1 - 1/e)$ approximation for monotone submodular maximization under cardinality constraints~\cite{nemhauser1978analysis}.
Facility location~\cite{cornuejols1977uncapacitated} and determinantal point processes~\cite{kulesza2012determinantal} provide related frameworks for diverse subset selection.
\mmr leverages these connections for theoretical grounding while remaining practically simple.
